Introductory statistics with `intRo`

### 2016-02-05 jcgs-paper-2016/paper.Rmd
Paper edits

---
title: "Introducing statistics with `intRo`"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
abstract: |
  `intRo` is a modern web-based application for performing basic data analysis and statistical routines. Leveraging the power of `R` and `Shiny`, `intRo` implements common statistical functions in a powerful and extensible modular structure, while remaining simple enough for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as `Excel` and `JMP`. `intRo` is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying `intRo`, as well as highlight some challenges and advantages of reactive programming.
keywords: R, Shiny, Statistics, Teaching, Interactive
bibliography: refs.bib
blind: false
---

@@

The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may be unfamiliar to the introductory statistics student. This issue may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors would tend to use a tool like \texttt{R} in their own work. Still, signficant compromises must be made when using software like \texttt{JMP}, including dealing with software licenses, unsupported desktop platforms, and non-ideal user interface decisions.

In teaching Introduction to Business Statistics at Iowa State University, we witnessed profound struggles by students attempting to practice the introductory concepts discussed in class using current software. Scrimshaw notes in his manuscript that "open-ended packages, like any others, may create obstacles to learning simply through their lack of user-friendliness in the sheer mechanics of operating them, rather than any intrinsic difficulty in the content...." [-@scrimshaw2001computers] In our own experience teaching, students' struggles were often directly related to the use of the software and not any sort of fundamental misunderstanding of the material, in agreement with Scrimshaw's finding.

@@

\texttt{intRo} is a web-based tool to accompany an introductory statistics class. It is meant to assist in the learning of statistics rather than act as a stand-alone deliverer of statistics education, with the intention of being used in conjunction with a guided class. An accompanying R package, titled \texttt{intRo} and available on GitHub, assists in the downloading, running, and deploying of \texttt{intRo} instances.

We have three fundamental philosophies that guided the creation of \texttt{intRo}. In particular, \texttt{intRo} is \emph{easy} to use and can be an \emph{exciting} part of learning statistics. Additionally, \texttt{intRo} is an \emph{extensible} tool, allowing for a user of \texttt{intRo} to tailor the tool for his or her own classroom needs.

## Easy

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{images/user_experience.pdf}
\caption{A typical user experience of generating a result in \texttt{intRo}. In this instance, a mosaic plot (and its corresponding code) was created and saved to the final compendium.}
\label{fig:user_experience}
\end{figure}

In order for an application to be useful to students in their first statistics course, it can't hog the spotlight. Statistics as a subject should shine through as a fun and interesting topic to study, rather than being bogged down by difficult to learn software. With this in mind, in the development of \texttt{intRo} we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find exactly what they need easily. With this easy to use framework, we also made the functionality available in \texttt{intRo} the minimal necessary for an introductory statistics course. Additional functionality is nice, but if it clutters the UI beyond the point of usability, it will frustrate an introductory student. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in \texttt{intRo}, view the code producing the result, and save the results for later.

Beyond being simple, \texttt{intRo} is also consistent. The tool is organized around specific tasks a student may perform in the process of a data analysis, called modules. Each module maintains a consistent layout, helping the user to become familiar with the location of the options, the results, and the code. In figure \ref{fig:ui}, we highlight the five elements that comprise the \texttt{intRo} interface.

@@

# In the classroom
\texttt{intRo}'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as \texttt{JMP} and \texttt{Excel}, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. \texttt{JMP}, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using \texttt{R} itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like \texttt{rmarkdown}, but this again is a bit of a learning curve for the new statistics student).

# Using intRo
\texttt{intRo} can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package \texttt{intRo}. Currently, the package is only available and GitHub, and can be installed using the \texttt{devtools} package as follows:

@@

# Details

>>>>

---
title: "Introductory statistics with `intRo`"
author:
  - name: Eric Hare
    affiliation: Iowa State University
    department: Department of Statistics
  - name: Andee Kaplan
    affiliation: Iowa State University
    department: Department of Statistics
output:
  pdf_document:
    fig_caption: yes
    includes:
      in_header: preamble.tex
    number_sections: yes
    template: jcgs_article.tex
fontsize: 12pt
abstract: |
  `intRo` is a web-based application for performing basic data analysis and statistical routines. Leveraging the power of `R` and `Shiny`, `intRo` implements common statistical functions in an extensible modular structure, while including a point-and-click interface for the novice statistician. This simplicity lends itself to a natural presentation in an introductory statistics course as a substitute for other commonly used statistical software packages, such as `Excel` and `JMP`. `intRo` is currently deployed at the URL \url{http://www.intro-stats.com}. Within this paper, we introduce the application and explore the design decisions underlying `intRo`, as well as highlight some challenges and advantages of reactive programming.
keywords: Graphical User Interface, Interactivity, Reproducibility, Reactive Programming, Statistical Software, Programming Paradigms
bibliography: refs.bib
blind: false
---

@@

The widespread adoption of \texttt{R} as a tool for statistical analysis has undoubtedly been an important development for the scientific community. However, using \texttt{R} in most cases still requires a basic knowledge of programming concepts which may pose a steep learning curve for the introductory statistics student. This additional time commitment may explain why introductory courses often utilize software applications such as \texttt{JMP}, even though instructors tend to use a tool like \texttt{R} in their own work. This brings along signficant compromises must be made when using software like \texttt{JMP}, including dealing with software licenses and unsupported desktop platforms.

In teaching Introduction to Business Statistics at Iowa State University, we witnessed profound struggles by students attempting to practice introductory concepts discussed in class using current software. Scrimshaw [-@scrimshaw2001computers] notes in his manuscript that "open-ended packages, like any others, may create obstacles to learning simply through their lack of user-friendliness in the sheer mechanics of operating them, rather than any intrinsic difficulty in the content...." In our own experience teaching, students' struggles were often directly related to the use of the software and not any sort of fundamental misunderstanding of the material, in agreement with Scrimshaw's finding.

@@

\texttt{intRo} is a web-based tool to accompany an introductory statistics class. It is meant to assist in the learning of statistics rather than as a stand-alone delivery of statistics education, with the intention of being used in conjunction with a guided class. An accompanying R package, titled \texttt{intRo} and available on GitHub, assists in the downloading, running, and deploying of \texttt{intRo} instances.

Three fundamental philosophies that guided the creation of \texttt{intRo}. In particular, \texttt{intRo} is \emph{easy} to use and can be an \emph{exciting} part of learning statistics. Additionally, \texttt{intRo} is an \emph{extensible} tool, allowing for a user of \texttt{intRo} to tailor the tool for his or her own classroom needs.

In the development of \texttt{intRo}, we focused on aspects of the user interface (UI) and output that make it easy to pick up without extensive training. We used large, easy to click icons in the page header to help students find what they need more easily. We also made the functionality available the minimal necessary for an introductory statistics course. Figure \ref{fig:user_experience} illustrates the simple steps a student takes to generate a result in \texttt{intRo}. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.

\begin{figure}[H]
\centering
\includegraphics[width=\linewidth]{images/user_experience.pdf}
\caption{A typical user experience of generating a result in \texttt{intRo}. In this instance, a user clicks on the Graphical tab to create a mosaic plot. The user sees the plot, and elects to click the save button to store the plot (and its corresponding code) to the final compendium.}
\label{fig:user_experience}
\end{figure}

Beyond being simple, \texttt{intRo} is also consistent. The tool is organized around specific tasks a student may perform in the process of a data analysis, called modules. To the user, a module is simply a page of statistics functionality that maintains a consistent layout, helping the user to become familiar with the location of the options, the results, and the code. Figure \ref{fig:ui} highlights the five elements that comprise the \texttt{intRo} interface.

@@

# Using intRo

## intRo Package

\texttt{intRo} can be downloaded, ran, and deployed on ShinyApps.io through the use of the R package \texttt{intRo}. Currently, the package is only available and GitHub, and can be installed using the \texttt{devtools} package as follows:

@@

## In the classroom

\texttt{intRo}'s ease of use lends itself to a natural presentation in an introductory statistics course. Statistics software typically used in introductory courses, such as \texttt{JMP} and \texttt{Excel}, have restrictive software licensing and platform requirements that can limit the software's accessibility to students. Even if these limitations can be overcome, these software programs have non-ideal interfaces for an exploratory data analysis. \texttt{JMP}, for instance, hides several basic regression functions such as a line of best fit behind small red arrows. Using \texttt{R} itself in the classroom is an appealing alternative, but suffers from some other problems. Many students in introductory statistics courses have not been exposed to programming languages and concepts, which could require a substantial overhead in order to begin a basic statistical analysis. When results need to be submitted as homework, getting the output into a document can be a frustrating and tedious experience (this experience is made much easier with new tools like \texttt{rmarkdown}, but this again is a bit of a learning curve for the new statistics student).

# Understand intRo
